Unveiling Robustness Threats by Mining the Exception Stack Traces of over 600 Android Projects

### 2014-09-03 android-stacks.tex
all over

%GitHub 315,92
% Google Code: 127,456
%All: 159,048

@@

%the mobile industry shipped over 315 million smartphones in the last quarter of 2013
%The pervasiveness of the World Wide Web and the astonishing increase 
%in the number of smartphones stimulated the creation of applications which extend the phones
% capabilities far beyond of the basic calls and textual messages. 
%f we look at global smartphone shipments the numbers are astonishng;
%the mobile industry shipped over 315 million smartphones in the last quarter of 2013 [ref-googleIO].
%Such increase on smartphones combined with the pervasiveness of the World Wide Web
%have stimulated the creation of applications which extend the phones caVALIAR MOVER ESTE STACK TRACES PARA UMA FIGURA QUE MOSTR COMO ELE QUEBRADO... PELO EXCEPTION Mpabilities 
%far beyond of the basic calls and textual messages.

In this context, Android has become a leading platform - thanks to its open-source 
model and hardware partners like Samsung, HTC, Motorola, Asus (besides Nexus Google's
 own device) ~\cite{gartner}.  Google estimates that there is over 1 billion Android active users per month ~\cite{googleio},
and the number of apps built on top of Android increases in a daily rate.

@@

By convention an Error represents an unrecoverable condition which usually results
from failures detected by the Java Virtual Machine, such as OutOfMemoryError and
normally should not be handled inside the application. The user-defined exceptions 
can be either checked or runtime. Although the Java specification~\cite{gosling2000java} 
suggests that user-defined exceptions should be checked exception (because doing so 
the callers of a method will know about the exceptions that a method can throw and so 
they can decide what to do about them) there is a long lasting debate about the use of 
checked and unchecked exceptions in Java.

@@

The runtime exceptions, however are not only used to represent user-defined
exceptional conditions, they are used by the JVM when a program violates 
the semantic constraints of the Java programming language (out-of-bounds array index, divide-by-zero 
error, null pointer references) from which most of the programs are not expected to recover. 
\footnote{Some programming languages react to such errors by peremptorily terminating the program; 
other programming languages allow similar approaches e.g.C\#; and others let the program to continue
 its execution in some situations such as the out-of-bounds array index e.g., C++. }

@@

In Java, once an exception is thrown, the JVM looks for the nearest enclosing exception handler
(in Java it is represented by the try-catch block), and unwinds the execution stack if necessary.
 To look for the handler on the invocation stack is claimed to increase software reusability, 
since the invoker of an operation can handle it in a wider context ~/cite{miller1997issues}.

 A common way of  propagating exceptions in Java programs is through exception wrapping
 (also known as exception chaining). Exception wrapping allows one exception 
to be wrapped in another one and re-thrown. Figure~\ref{fig:wrapping} presents 
an exception stack trace which illustrates an exception wrapping. 
For simplicity we will refer to exception stack trace as stack trace along the text.
The bottom part of the stack trace is the \emph{root cause}, which indicates the
first reason for the error to be thrown (in this case, the computer run out of
memory). The top part of the stack trace indicates the location of the exception
manifestation (which will call \emph{top level wrapper} along this paper). The
execution flow  between the root cause and the exception manifestation may
include other intermediate exception wrappings. In all levels, the exception
\emph{signaler}, is the method that threw the exception, represented on the
stack trace as the first method call below the exception declaration.

@@

There is a long-lasting debate about the pros and cons of checked and unchecked exceptions 
~\cite{javatut,stackoverlow,debate},\footnote{152 questions in Stackoverlow
are related to this debate}. Such debates, however, are mainly based on the developers own experience in
using both approaches - since there is a lack of empirical studies on their the pros and cons.

Nevertheless, several general guidelines have been proposed on how to use Java
exceptions~\cite{mandrioli1992advances,gosling2000java,wirfs2006toward,
bloch2008effective}. Such guidelines do not focus on 
advocating any specific exception type, but rather propose ways to effectively use each of them.
 We compiled a list of the guidelines related to exception types in Java, 
summarized below:\footnote{We also compiled guidelines related
to exception handling but they are out of the scope of this paper.}

@@

\emph{IV- All exceptions explicitly thrown by a method should be documented.}
(\cite{mandrioli1992advances,gosling2000java,wirfs2006toward,bloch2008effective}).
The exceptions thrown by a method are an important part of methods interface,
and is required to use the method properly. The checked exceptions are already
part of the  methods signature, and the method caller is aware of the checked
exceptions being thrown by it. According to~\cite{bloch2008effective}, it is
also wise to document the explicitly thrown runtime exceptions\footnote{This
excludes the implicitly signaled JVM runtime exceptions related to programming
mistakes} as carefully as checked exceptions. Doing so, the clients of a method
will be aware of all the exceptions the method can throw and can protect its code
against unforeseen exceptions.

If the developer fails to do follow such practices (specially when developing 
library/reusable code) it will be difficult or even impossible for the caller to 
make effective use of such method~\cite{wirfs2006toward, bloch2008effective},
and protect the system against the uncaught exceptions - one of the main causes of 
systems crashes.

These best practices guided our exploratory study described next.
Based on information extracted from stack traces (complemented by
bytecode and source-code analysis), we investigate whether stack characteristics
can reveal whether these practices have been obeyed.

@@

\section{Empirical Study Design}
\label{sec:study}

@@

Our study focused on open-source apps, since the information needed to perform our study
cannot be retrieved from commercial apps, whose issue reports and 
source code are not publicly available. The Android open-source apps were also 
target of other research studies ~\cite{Linar13,ahimed}.

@@

  \item[RQ1] \noindent\emph{What are the common characteristics of exception stack traces reported on issues?} 
Some of the common characteristics to be investigated are the exceptions reported as the root cause on
the exception stack trace, the origin of such exceptions (e.g., the application, a library, the Android platform).

@@

%Which patterns of exception (mis)use emerge?
  \item[RQ2] \noindent\emph{Which fragilities on the exception-related code emerge from them?}
Based on a set of best practices compiled in this study, we investigated whether characteristics of exception stack traces  
can pinpoint best practices violations. Such  violations reveal fragilities
on the exception-related code, which threats the application dependability.

@@

@@

To answer the research questions the following information is needed: (i) the issues of reported on every Android project hosted on 
GitHub and Google Code; (ii) the stack traces embedded on those issues; (iii) the type of the exceptions
 reported on the stack traces (e.g., error, runtime, checked); (iv) the origin of such exceptions 
(e.g., the application, a library, the Android platform). Next sections describes how such information 
was retrieved from GitHub and Google Code, how the exceptions stack traces were extracted and distilled,
 and how the type associated to each exception and its origins was identified.

@@

This study used the dataset provided by the GHTorrent project~\cite{Gousi13}, 
an off-line mirror of the data  offered through the Github API.  
To identify Android projects, we performed a case insensitive search for the
term \textsf{android} in the repository's names and short descriptions.  
Up to 23 Feb 2014,  when we queried GHTorrent, such heuristic could
 find 2.542 repositories,  from which 589 apps had at least one
issue containing a stack trace.

Then we performed further cleanup, inspecting the site of every Android
reporting at least one stack trace, to make sure that they represented real
mobile apps. During this clean up 106 apps were removed because they were either
example projects (i.e., toy projects) or tools to support Android development
(e.g. Selendroid, Roboeletric - tools to support the testing of Android apps).
The filtered set consisted of 482 apps.

%4,208

This set of 482 projects contained overall 31,592 issues from which 4,042 exception stack traces 
were extracted. Issues on Github are different from issues on dedicated bug tracking tools such as 
Bugzilla and Jira. The most important difference is that there are no predefined fields
  (e.g. severity and priority). Instead, Github uses a more open ended tagging system, on which
repositories are offered a pre-defined set of labels, but repository owners can modify 
them at will. Therefore, an issue may have none ore an arbitrary set of labels depending 
on which repository it was created. Table ~\ref{tab:lables} illustrates the ocurrences of different lables 
on the issues including exception stack traces.

@@

Moreover, issues and pull requests are dual on Github; all pull requests have a corresponding 
``backing'' issue which are automatically generated. Therefore, we excluded such automatically generated
issues from our analysis. Finally, the Github API allows the automated
generation of issues for specific repositories, which automated tools often use
to report crashes. In some cases, this led to high number of issues that
included stack traces. We identified and filtered those cases out (e.g.,
the~\textsf{pullwifi} project was responsible for almost 50\% of all Android issues in our dataset).

@@

However, Google Code still contain largely used open-source Android apps (e.g. K9Mail \footnote{K9Mail moved to Github but as a way of not loosing the project history 
it advices their users to report bugs on Google Code issue tracker.})  and the history of such apps are also valuable. 
To overcome this limitation we needed to implement a Web Crawler (incorporated in Exception Miner tool described next) that navigated 
through the web interface of Google Code projects extracting all issues and issue comments and storing in a relational database for later analysis.

To identify Android projects in Google Code, we performed a similar heuristic, we performed a case insensitive search 
(on Google Code search interface) for the term \textsf{android}.  On January 2014, when we queried Google Code, such heuristic could
 find 788  projects, from which 724 defined at least 1 issue and, 183 projects 
had at least one issue including a stack trace. Then we performed further cleanup, inspecting the site of every Android
reporting at least one stack trace, and as a result we could identify 157. 
Android projects.

This set of 157 projects contained overall 127,456 issues from which 1,963 exception stack traces 
were extracted. Based on the same assumption used for Github dataset selection, we did not retrict 
the analysis for issues labled as Defect.

>>>>

%GitHub 31,592
% Google Code: 127,456
%All: 159,048
% bf: 31,592 (with exception)

@@

%the mobile industry shipped over 315 million smartphones in the last quarter of 2013
%The pervasiveness of the World Wide Web and the astonishing increase 
%in the number of smartphones stimulated the creation of applications which extend the phones
% capabilities far beyond of the basic calls and textual messages. 
%f we look at global smartphone shipments the numbers are astonishng;
%the mobile industry shipped over 315 million smartphones in the last quarter of 2013 [ref-googleIO].
%Such increase on smartphones combined with the pervasiveness of the World Wide Web
%have stimulated the creation of applications which extend the phones caVALIAR MOVER ESTE STACK TRACES PARA UMA FIGURA QUE MOSTR COMO ELE QUEBRADO... PELO EXCEPTION Mpabilities 
%far beyond of the basic calls and textual messages. 
315,92
In this context, Android has become a leading platform - thanks to its open-source 
model and hardware partners like Samsung, HTC, Motorola, Asus (besides Nexus Google's
 own device) ~\cite{gartner}.  Google estimates that there is over 1 billion Android active users per month ~\cite{googleio},
and the number of apps built on top of Android increases in a daily rate.

@@

By convention an Error represents an unrecoverable condition which usually results
from failures detected by the Java Virtual Machine, such as OutOfMemoryError and
normally should not be handled inside the application. The user-defined exceptions 
can be either checked or runtime. The Java specification~\cite{gosling2000java} 
suggests that user-defined exceptions should be checked (because doing so 
the callers of a method will know about the exceptions that a method can throw and so 
they can decide what to do about them). However, runtime exceptions can also be 
used to represent user-defined exceptional conditions.In fact, there is a long-lasting debate 
about the pros and cons of both approaches ~\cite{javatut,stackoverlow,debate},
\footnote{152 questions in Stackoverlow are related to this debate}.

@@

Besides, the runtime exceptions are also implicitly thrown by the JVM when a program violates 
the semantic constraints of the Java programming language (e.g., out-of-bounds array index, divide-by-zero 
error, null pointer references).\footnote{Some programming languages react to such errors by peremptorily terminating the program; 
other programming languages allow similar approaches e.g.C\#; and others let the program to continue
 its execution in some situations such as the out-of-bounds array index e.g., C++. }  
The programs are not expected to handle such JVM runtime exceptions ~\cite{gosling2000java} s.

@@

In Java, once an exception is thrown, the JVM looks for the nearest enclosing exception handler
(in Java it is represented by the try-catch block), and unwinds the execution stack if necessary.
 Looking for the handler on the invocation stack is claimed to increase software reusability, 
since the invoker of an operation can handle it in a wider context ~/cite{miller1997issues}.

 A common way of  propagating exceptions in Java programs is through exception wrapping
 (also known as exception chaining). Exception wrapping allows one exception 
to be wrapped in another one and re-thrown. Figure~\ref{fig:wrapping} shows 
an exception stack trace which illustrates an exception wrapping. 
For simplicity we will refer to exception stack trace as stack trace along the text.
The bottom part of the stack trace is the \emph{root exception}, which indicates
the first reason (root cause) for the exception be thrown (in this case, the computer run out of
memory). The top part of the stack trace indicates the location of the exception
manifestation (which will call \emph{exception wrapper} along this paper). The
execution flow  between the root exception and the wrapper may
include other intermediate exception wrappers. In all levels, the exception
\emph{signaler}, is the method that threw the exception, represented on the
stack trace as the first method call below the exception declaration.

@@

Several general guidelines have been proposed on how to use Java
exceptions~\cite{mandrioli1992advances,gosling2000java,wirfs2006toward,
bloch2008effective}. Such guidelines do not focus on 
advocating any specific exception type, but rather propose ways to effectively use each of them.
 We compiled a list of the guidelines related to exception types in Java, 
summarized below:\footnote{We also compiled guidelines related
to exception handling but they are out of the scope of this paper.}

@@

\emph{IV- All exceptions explicitly thrown by a method should be documented.}
(\cite{mandrioli1992advances,gosling2000java,wirfs2006toward,bloch2008effective}).
The exceptions thrown by a method are an important part of methods interface,
and is required to use the method properly. The checked exceptions are already
part of the  methods signature, and the method caller is aware of the checked
exceptions being thrown by it. According to~\cite{bloch2008effective}, it is
also wise to document the explicitly thrown runtime exceptions\footnote{This
excludes the implicitly signaled JVM runtime exceptions related to programming
mistakes} as carefully as checked exceptions. Doing so, the clients of a method
will be aware of all the exceptions the method can throw and can protect design the code
to deal with them ~\cite{Robil00}.

If the developer fails to do follow such practices (specially when developing 
utility code) it will be difficult or even impossible for the caller to 
make effective use of such method~\cite{wirfs2006toward, bloch2008effective},
and protect the system against unforeseen exceptions thrown by the utility code,
which can lead to the uncaught exceptions - one of the main causes of 
systems crashes. These best practices guided our exploratory study dscribed next.

@@

\section{Study Design}
\label{sec:study}

@@

Our study focused on open-source apps, since the information needed to perform our study
cannot be retrieved from commercial apps, whose issue report systems and 
source codes are not publicly available. The Android open-source apps were also 
target of other research studies ~\cite{Linar13,ahimed}.

@@

  \item[RQ1] \noindent\emph{What are the common characteristics of exception stack traces reported on issues?}

@@

     \item

%Which patterns of exception (mis)use emerge?
  \item[RQ2] \noindent\emph{Which fragilities on the exception-related code emerge from them?}

@@

Some of the common characteristics to be investigated are the exceptions reported as the root cause on
the exception stack trace, the origin of such exceptions (e.g., the application, a library, the Android platform). 
Based on a set of best practices compiled in this study, we investigated 
whether characteristics of exception stack traces can pinpoint best practices violations. 
Such  violations reveal fragilities
on the exception-related code, which threats the application dependability.

@@

To answer the research questions the following information is needed: (i) the issues reported on Android project hosted on 
GitHub and Google Code; (ii) the stack traces embedded on these issues; (iii) the type of the exceptions
 reported on the stack traces (e.g., error, runtime, checked); (iv) the origin of such exceptions 
(e.g., the application, a library, the Android platform). Next sections describe: how such information 
was retrieved from GitHub and Google Code; how the exceptions stack traces were extracted and distilled,
 and how the type associated to each exception and its origins was identified, and finally how manual
inspections supported this work.

@@

This study used the dataset provided by the GHTorrent project~\cite{Gousi13}, 
an off-line mirror of the data  offered through the Github API.  
To identify Android projects, we performed a case insensitive search for the
term \textsf{android} in the repository's names and short descriptions.  
Up to 23 Feb 2014,  when we queried GHTorrent, such heuristic could
 find 2.542 repositories. Running the ExceptionMiner tool (described next)
 in this set we observed that 589 apps had at least one issue containing a stack trace.

Then we performed further cleanup, inspecting the site of every Android
reporting at least one stack trace, to make sure that they represented real
mobile apps. During this clean up 106 apps were removed because they were either
example projects (i.e., toy projects) or tools to support Android development
(e.g. Selendroid, Roboeletric - tools to support the testing of Android apps).
The filtered set consisted of 482 apps. This set of 482 projects contained overall 31,592 issues from which 4,042 exception stack traces 
were extracted. Issues on Github are different from issues on dedicated bug tracking tools such as 
Bugzilla and Jira. The most important difference is that there are no predefined fields
  (e.g. severity and priority). Instead, Github uses a more open ended tagging system, on which
repositories are offered a pre-defined set of labels, but repository owners can modify 
them at will. Therefore, an issue may have none ore an arbitrary set of labels depending 
on which repository it was created. Table ~\ref{tab:lables} illustrates the ocurrences of different lables 
on the issues including exception stack traces.

@@

Moreover, issues and pull requests are dual on Github; all pull requests have a corresponding 
``backing'' issue which are automatically generated. Therefore, we excluded such automatically generated
issues from our analysis.

@@

However, Google Code still contains largely used open-source Android apps (e.g. K9Mail \footnote{K9Mail moved to Github but as a way of not loosing the project history 
it advices their users to report bugs on Google Code issue tracker.})  and the history of such apps are also valuable. 
To overcome this limitation we needed to implement a Web Crawler (incorporated in Exception Miner tool described next) that navigated 
through the web interface of Google Code projects extracting all issues and issue comments and storing in a relational database for later analysis.

To identify Android projects in Google Code, we performed a similar heuristic: we performed a case insensitive search 
(on Google Code search interface) for the term \textsf{android}.  On January 2014, when we queried Google Code, such heuristic could
 find 788  projects, from which 724 defined at least 1 issue and, 183 projects 
had at least one issue including a stack trace. Then we performed further cleanup, inspecting the site of each project,
 and as a result we could identify 157 Android projects.

This set of 157 projects contained overall 127,456 issues, from which 1,963 exception stack traces 
were extracted. Table ~\ref{tab:lables} illustrates the ocurrences of different lables 
on the issues including exception stack traces. Doferently from GitHib in Google Code most of 
the lables were labled as ``Defect" . On the study we considered all issues reporting stacks (regardless its lables)
based on the same assumption described for the GitHub repository.
